Runtime-Assisted Convergence in Replicated Data Types

### 2022-02-09 pldi22_rebuttal.md

We thank the reviewers for their valuable feedback. We address major
concerns first, followed by reviewer-specific comments.

Use of Synchronization
----------------------

Merges in Carmot use synchronization. However, synchronization does
not trivially guarantee convergence. Our contribution is to show
*how* to use synchronization in a way that (a). Replicas are always
available for user actions, (b). No user action has to wait for a
remote replica to respond, and (c). Replicas eventually converge. To
demonstrate the subtlety in our use of synchronization, consider the
following trivial synchronizations:

1. Each operation is synchronized -- the system would violate
   aforementioned properties (a) and (b).
2. Merges ``stop the world'' to pull the latest version from all the
   branches -- violates properties (a) and (b) as no replica can
   progress when one replica is merging.
3. Only merges are synchronized; each merge is aware of, and has all
   the information from, the previous merges -- this would linearize
   merges *in time*, but nonethless results in ill-formed histories
   that violate property (c). Fig e.g., in Fig.5(b) merges could be
   ordered linearly in time, yet result in divergence.

Our key discovery is the LCA-based premise of Merge and FastFwd
rules that effectively ensure that LCAs in any history are organized
as a lattice. This inturn is based on our observation that multiple
LCAs inevitably lead to divergence in general case. We use
synchronization to enable the checks on LCAs and enforce the lattice
property (Corollary 3.6). Otherwise Synchronization in itself doesn't
constitute the solution.

Comparison with MRDTs and CRDTs
--------------------------------

Despite their best efforts to ensure convergence, Kaki et al's MRDT
merges [13] nonetheless result in divergence. For e.g., our
`Set.merge` in Sec.1 is same as Kaki et al's set merge, yet it results
in divergence as shown in Figs.4(b) and 5(b). The existence of these
counterexamples is in fact why they were not able to prove MRDT
convergence (source: personal communication). While convergence is
only a small aspect of full-functional correctness, ensuring it in the
context of CRDTs and MRDTs remains extremely challenging even for the
experts
([here](https://twitter.com/martinkl/status/1327025979454263297?lang=en)
is a recent example from CRDTs).  We hope our formal guarantees of
convergence would make RDTs accessible to non-experts.

Evaluation
------------

We have chosen to do one in-depth case study instead of experimenting
with a collection of RDTs as it lets us examine the tradeoffs more
closely. All the MRDT benchmarks from Kaki et al continue to work on
Carmot with no increase in latency and diff size measurements. Carmot
is in fact a fork of Quark -- the system from Kaki et al. Carmot
extends Quark with an more efficient network and storage layer (due to
Scylla), and a runtime component that is activated only during a merge
(happens in a background thread). **There is no additional overhead
associated with a user action**. When a user submits an action, it is
immediately executed and committed to the local version. Thus, insofar
as user interaction is concerned, Carmot is qualitatively no different
from any other CRDT system. Unfortunately, we couldn't find a
standardized CRDT system (akin to Quark for MRDTs) for us to do
comparisons against.

It is possible to give theoretical upper bound on staleness in
Carmot. Ring topology guarantees the uniqueness of LCAs (Lemma 3.4) by
construction (without a need for synchronization). If replicas are
organized as a ring, the maximum staleness is the sum of latencies on
the circumference of the ring. Hence if we have replicas distributed
across US-east and US-west, it is reasonable to expect staleness to be
in the order of 100ms. As it stands now, our implementation simply
reflects our formal model, which includes no such optimizations.
Lastly, we'd like to note that mere presence of synchronization doesn't
necessarily lead to unacceptable performance. Google docs, for
instance, relies on a centralized OT server to coordinate all its
edits and yet has acceptable performance.

Reviewer A
----------

* The concrete semantics of Sec.4 *refines* the abstract semantics of
  Sec. 3. The proof we did in Ivy is in fact a refinement proof (can be
  found in `quarkRefineImplement.ivy` in the supplementary material). We
  are incorrect in stating that all properties have been re-proved for
  the concrete semantics. The fact is that the properties follow
  because Sec.4 semantics is a refinement of Sec.3 semantics. We
  apologize for the confusion.

* For comments on Related Work and Evaluation, please see above.

Reviewer B
----------

* Useful merge functions exist for more complicated data structures.
  In [13] Kaki et al provide a constructive definition of well-formed
  merges with help of *relational decomposition*. Albeit
  computationally inefficient, relational decomposition leads to
  simple yet useful merges for a range of data structures. The
  problem, however, is that [13] lacks either a system or a proof that
  guarantees the convergence of resultant MRDTs. As demonstrated
  above, divergence is possible despite a sensible merge semantics.
  Carmot closes this gap.

Reviewer C
----------

* We hope your concerns on staleness and performance are addressed by
  our note on Evaluation above.

Reviewer D
-----------

* Every client forks off a branch which is effectively a replica, so
  there are multiple replicas per machine. With 30 clients, there are
  30 branches with 30 replicas distributed across 3 machines.

* Inclusion of a runtime imposes no additional overhead on the latency
  of individual operations. Given that Carmot runtime uses
  synchronization for merges, and the overhead of synchronization
  increases at least linearly with the number of replicas, one would
  expect the latency of each operation to grow at least linearly if it
  is affected by the presence of runtime. However, as Fig.10
  demonstrates, that's not the case. Hence the conclusion that our
  approach imposes no additional cost on latency and availability (our
  latency measurement accounts for availability too as it measures the
  time from the operation is submitted to the time a result is
  received. If the availability of the system goes down, then
  operations have to wait longer to find a replica, increasing their
  overall latency. Fig.10 shows this is not the case with Carmot.)

* We hope your other comments on performance are addressed by our note
  on Evaluation above.

### 2022-02-09 pldi22_rebuttal.md

We thank the reviewers for their valuable feedback. We address major
concerns first, followed by reviewer-specific comments.

Use of Synchronization
----------------------

While merges in Carmot use synchronization, it is only one aspect of
our overall approach. Our contribution is to show *how* to use
synchronization in a way that (a). Replicas are always available for
user actions, (b). No user action has to wait for a remote replica to
respond, and (c). Replicas eventually converge. There are several ways
one could use synchronization yet fail to achieve the aforementioned
objectives. For e.g:

1. Each user operation is synchronized: this leads to violation of
   properties (a) and (b).
2. Merges synchronize by ``stoping the world'' to pull the latest
   version from all the branches. This too violates properties (a) and
   (b) as no replica can progress when one replica is merging.
3. Merges are synchronized with each other; each merge is aware of,
   and has all the information from, the previous merges: this would
   linearize merges *in time*, but nonethless results in ill-formed
   histories that violate property (c). Fig e.g., in Fig.5(b) merges
   could be ordered linearly in time, yet result in divergence.

As summarized by Reviewer B, our key discovery is the well-formedness
condition that translates to LCA-based premises of Merge and FastFwd
rules. These conditions effectively ensure that LCAs in any history
DAG are organized as a lattice, which gives us the unique LCA property
(Lemma 3.4), and thereby convergence (Theorem 3.7). The reasoning
about LCAs is critical to convergence, and cannot be done away with.
We use synchronization to perform the the checks on LCAs, which inturn
guarantees convergence. Otherwise synchronization in itself doesn't
guarantee convergence.

@@

Comparison with MRDTs and CRDTs
--------------------------------

Despite their best efforts to ensure convergence, Kaki et al's MRDT
merges [13] nonetheless result in divergence. For e.g., our
`Set.merge` in Sec.1 is same as Kaki et al's set merge, yet it results
in divergence as shown in Figs.4(b) and 5(b). The existence of these
counterexamples is in fact why they were not able to prove MRDT
convergence (source: personal communication). While convergence is
only a small aspect of full-functional correctness, getting it right
is not easy, even for the experts
([Here](https://twitter.com/martinkl/status/1327025979454263297?lang=en)
is a recent example from CRDTs. Another example is the queue
persistent data structure from [Farinier et
al](https://hal.inria.fr/hal-01099136v1/document), which turned out to
admit the multiple LCA anomaly and (thereby) divergence). We hope
our formal guarantees help the community advance the discussion on
RDTs past convergence.

@@

Evaluation
------------

We have chosen to do one in-depth case study instead of experimenting
with a collection of RDTs as it lets us examine the tradeoffs more
closely. All the MRDT benchmarks from Kaki et al continue to work on
Carmot with no increase in latency and diff size measurements. Carmot
is in fact a fork of Quark -- the system from Kaki et al. Carmot
extends Quark with an more efficient network and storage layer (due to
Scylla), and a runtime component that is activated only during a merge
(happens in a background thread). **It is important to note that
Carmot does not introduce any additional overhead on a user action to
guarantee convergence**. When a user submits an action, it is
immediately executed and committed to the local version. Thus, insofar
as user interaction is concerned, Carmot is qualitatively no different
from any other CRDT system. Unfortunately, we couldn't find a
standardized CRDT system (akin to Quark for MRDTs) for us to do
comparisons against.

It is possible to give theoretical upper bound on staleness in Carmot.
We have proved (elsewhere) that a ring topology guarantees the
uniqueness of LCAs by construction without a need for global
synchronization. If replicas are organized as a ring, the maximum
staleness is the sum of latencies on the circumference of the ring.
Hence if we have replicas distributed across US-east and US-west, it
is reasonable to expect staleness to be in the order of 100ms. As it
stands now, our implementation simply reflects our formal model, which
includes no such optimizations.  Lastly, we'd like to note that there
exist real distributed applications that make use of synchronization
in practice. Google docs, for instance, relies on a centralized OT
server to coordinate all its edits yet remains usable.

@@

* The concrete semantics of Sec.4 *refines* the abstract semantics of
  Sec. 3. The proof we did in Ivy is in fact a refinement proof (can be
  found in `quarkRefineImplement.ivy` in the supplementary material). We
  are incorrect in stating that all properties have been re-proved for
  the concrete semantics. The fact is that the properties follow
  because Sec.4 semantics is a refinement of Sec.3 semantics. We
  apologize for the confusion.

* For comments on Related Work and Evaluation, please see above.

Reviewer B
----------

* Useful merge functions exist for more complicated data structures.
  In [13] Kaki et al provide a constructive definition of well-formed
  merges with help of *relational decomposition*. Albeit
  computationally inefficient, relational decomposition leads to
  simple yet useful merges for a range of data structures. The
  problem, however, is that [13] lacks either a system or a proof that
  guarantees the convergence of resultant MRDTs. We show that
  divergence is possible despite a sensible merge semantics, and close
  this gap.

* Git acknowledges the possibility of multiple LCAs in its
  [documentation](https://git-scm.com/docs/merge-strategies). The
  default merge strategy of Git is `recursive`, which recursively
  merges LCAs if there are multiple of them. They admit that this is
  merely a heuristic. We have found a pathological case where such
  recursive merging could be a problem (Full example with
  illustrations included in the appendix). We haven't contacted the
  Git maintainers yet.

@@

* We hope your concerns on staleness and performance are addressed by
  our note on Evaluation above.

Reviewer D
-----------

* Every client forks off a branch which is effectively a replica, so
  there are multiple replicas per machine. With 30 clients, there are
  30 branches with 30 replicas distributed across 3 machines.

* Inclusion of a runtime imposes no additional overhead on the latency
  of individual operations. Given that Carmot uses synchronization for
  merges, and the overhead of synchronization increases at least
  linearly with the number of replicas, one would expect the latency
  of each operation to grow at least linearly *if* the runtime
  interferes with the normal execution. Fig.10 categorically
  demonstrates that this is not the case. Hence the conclusion that our
  approach imposes no additional cost on latency and availability (our
  latency measurement accounts for availability too as it measures the
  time from the operation is submitted to the time a result is
  received. If the availability of the system goes down, then
  operations have to wait longer to find a replica, increasing their
  overall latency. Fig.10 shows this is not the case with Carmot.)

* Being based on MRDTs, our approach *is* indeed comparable to CRDTs.
  The absence of synchronization in CRDTs is a result of careful
  engineering by domain experts [20, 25, 26, 11]. In the absence of
  such expertise, the only alternative available to an application
  developer is to synchronize every operation. Our approach presents a
  third alternative -- a compromise where developer need not
  synchronize every operation, yet obtain convergence.

* Our claim in contribution bullet #2 pertains to the formalization of
  the abstract machine and the theorems we prove on it. Mechanization
  in Ivy, similar to mechanization in Coq, is intended to convey the
  confidence in the soundness of our meta-theory. Our Ivy development
  very closely resembles our formalization in Sections 3 and 4
  (including the theorems). Lemmas 3.4 to 3.8 have been proved in Ivy.
  In addition, we also proved that the concrete semantics of Sec. 4
  refine the abstract semantics of Sec. 3. While Sledgehammer extends
  an interactive proof assistant (Isabelle/HOL) with SMT support, our
  mechanized verification in Ivy is wholly SMT-driven; there is no
  interactive theorem proving involved (we have separately done manual
  proofs however). We shall include more details on Ivy formalization,
  and the comparison with Sledgehammer approach in the final version.

* We hope your questions on merge synchronization are answered by the
  note on Evaluation above.

### 2022-02-09 pldi22_rebuttal.md

Use of Synchronization
----------------------

While merges in Carmot use synchronization, it is only one aspect of
our overall approach. Our contribution is to show *how* to use
synchronization in a way that (a). Replicas are always available for
user actions, (b). No user action has to wait for a remote replica to
respond, and (c). Replicas eventually converge. There are several ways
one could use synchronization yet fail to achieve the aforementioned
objectives. For e.g:

1. Each user operation is synchronized: this leads to violation of
   properties (a) and (b).
2. Merges synchronize by ``stoping the world'' to pull the latest
   version from all the branches. This too violates properties (a) and
   (b) as no replica can progress when one replica is merging.
3. Merges are synchronized with each other; each merge is aware of,
   and has all the information from, the previous merges: this would
   linearize merges *in time*, but nonethless results in ill-formed
   histories that violate property (c). Fig e.g., in Fig.5(b) merges
   could be ordered linearly in time, yet result in divergence.

As summarized by Reviewer B, our key discovery is the well-formedness
condition that translates to LCA-based premises of Merge and FastFwd
rules. These conditions effectively ensure that LCAs in any history
DAG are organized as a lattice, which gives us the unique LCA property
(Lemma 3.4), and thereby convergence (Theorem 3.7). The reasoning
about LCAs is critical to convergence, and cannot be done away with.
We use synchronization to perform the the checks on LCAs, which inturn
guarantees convergence. Otherwise synchronization in itself is not
very useful.

Conversely, synchronization is not necessarily an antithesis to
practical adoption. Google docs, for instance, famously relies on a
centralized OT server to coordinate all its edits.

@@

Despite their best efforts to ensure convergence, Kaki et al's MRDT
merges [13] nonetheless result in divergence. For e.g., our
`Set.merge` in Sec.1 is same as Kaki et al's set merge, yet it results
in divergence as shown in Figs.4(b) and 5(b). The existence of these
counterexamples is in fact why they were not able to prove MRDT
convergence (source: personal communication). While convergence is
only a small aspect of full-functional correctness, getting it right
is not easy, even for the experts
([Here](https://twitter.com/martinkl/status/1327025979454263297?lang=en)
is a recent example from CRDTs. Another example is the queue
persistent data structure from [Farinier et
al](https://hal.inria.fr/hal-01099136v1/document), which turned out to
admit the multiple LCA anomaly and (thereby) divergence). We hope
our formal guarantees help the community advance the discussion on
RDTs past convergence.

Evaluation
------------

We have chosen to do one in-depth case study instead of experimenting
with a collection of RDTs as it lets us examine the tradeoffs more
closely. All the MRDT benchmarks from Kaki et al continue to work on
Carmot with no increase in latency and diff size measurements. Carmot
is in fact a fork of Quark -- the system from Kaki et al. Carmot
extends Quark with an more efficient network and storage layer (due to
Scylla), and a runtime component that is activated only during a merge
(happens in a background thread). **It is important to note that
Carmot does not introduce any additional overhead on a user action to
guarantee convergence**. When a user submits an action, it is
immediately executed and committed to the local version. Thus, insofar
as user interaction is concerned, Carmot is qualitatively no different
from any other CRDT system. Unfortunately, we couldn't find a
standardized CRDT system (akin to Quark for MRDTs) for us to do
comparisons against.

It is possible to give theoretical upper bound on staleness in Carmot.
We have proved (elsewhere) that a ring topology guarantees the
uniqueness of LCAs by construction without a need for global
synchronization. If replicas are organized as a ring, the maximum
staleness is the sum of latencies on the circumference of the ring.
Hence if we have replicas distributed across US-east and US-west, it
is reasonable to expect staleness to be in the order of 100ms. As it
stands now, our implementation simply reflects our formal model, which
includes no such optimizations.

Lastly, we'd like to point out that this work introduces staleness as
a novel tradeoff against convergence. While trading off staleness may
not be an optimal choice for every application, the fact that there
exists such a tradeoff is, in our opinion, noteworthy. And if an
application has to chose between latency and staleness, staleness is
arguably the less disruptive choice.

### 2022-02-09 pldi22_rebuttal.md

* The concrete semantics of Sec.4 *refines* the abstract semantics of
  Sec. 3. The proof we did in Ivy is in fact a refinement proof (can be
  found in `quarkRefineImplement.ivy` in the supplementary material). We
  are incorrect in stating that all properties have been re-proved for
  the concrete semantics. The fact is that the properties follow
  because Sec.4 semantics is a refinement of Sec.3 semantics. We
  apologize for the confusion.

* For comments on Related Work and Evaluation, please see above.

Reviewer B
----------

* Useful merge functions exist for more complicated data structures.
  In [13] Kaki et al provide a constructive definition of well-formed
  merges with help of *relational decomposition*. Albeit
  asymptotically sub-optimal, relational decomposition leads to simple
  yet useful merges for a range of data structures. The problem,
  however, is that [13] lacks either a system or a proof that
  guarantees the convergence of resultant MRDTs. We show that
  divergence is possible despite a sensible merge semantics, and close
  this gap.

* Git acknowledges the possibility of multiple LCAs in its
  [documentation](https://git-scm.com/docs/merge-strategies). The
  default merge strategy of Git is `recursive`, which recursively
  merges LCAs if there are multiple of them. This however merely a
  heuristic. We have found a pathological case where such recursive
  merging could become a problem (Full example with illustrations
  included in the appendix).
  [Others](http://r6.ca/blog/20110416T204742Z.html) have also found
  cases where Git merge is inconsistent. We haven't contacted the Git
  maintainers yet.

@@

Reviewer C
----------

* We hope the reviewer's concerns on staleness and performance are
  addressed by our note on Evaluation above.

### 2022-02-09 pldi22_rebuttal.md

We thank the reviewers for their valuable feedback. We address major
concerns first, followed by reviewer-specific comments.

Use of Synchronization
----------------------

While merges in Carmot use synchronization, it is only one aspect of
our overall approach. Our contribution is to show *how* to use
synchronization in a way that (a). Replicas are always available for
user actions, (b). No user action has to wait for a remote replica to
respond, and (c). Replicas eventually converge. There are several ways
one could use synchronization yet fail to achieve the aforementioned
objectives. For e.g:

1. Each user operation is synchronized: this leads to violation of
   properties (a) and (b).
2. Merges synchronize by ``stoping the world'' to pull the latest
   version from all the branches. This too violates properties (a) and
   (b) as no replica can progress when one replica is merging.
3. Merges are synchronized with each other; each merge is aware of,
   and has all the information from, the previous merges: this would
   linearize merges *in time*, but nonethless results in ill-formed
   histories that violate property (c). For e.g., in Fig.5(b) merges
   could be ordered linearly in time, yet result in divergence.

@@

Comparison with MRDTs and CRDTs
--------------------------------

Despite their best efforts to ensure convergence, Kaki et al's MRDT
merges [13] nonetheless result in divergence. For e.g., our
`Set.merge` in Sec.1 is same as Kaki et al's set merge, yet it results
in divergence as shown in Figs.4(b) and 5(b). The existence of these
counterexamples is in fact why they were not able to prove MRDT
convergence (source: personal communication). While convergence is
only a small aspect of full-functional correctness, getting it right
is not easy, even for the experts
([Here](https://twitter.com/martinkl/status/1327025979454263297?lang=en)
is a recent example from CRDTs. Another example is the queue
persistent data structure from [Farinier et
al](https://hal.inria.fr/hal-01099136v1/document), which turned out to
admit the multiple LCA anomaly and (thereby) divergence). We hope
our formal guarantees help the community advance the discussion on
RDTs past convergence.

Evaluation
------------

We have chosen to do one in-depth case study instead of experimenting
with a collection of RDTs as it lets us examine the tradeoffs more
closely. All the MRDT benchmarks from Kaki et al [13] continue to work
on Carmot with no increase in latency and diff size measurements.
Carmot is in fact a fork of Quark -- the system from Kaki et al.
Carmot extends Quark with an more efficient network and storage layer
(due to Scylla), and a runtime component that is activated only during
a merge (happens in a background thread). **It is important to note
that Carmot does not introduce any additional overhead on a user
action to guarantee convergence**. When a user submits an action, it
is immediately executed and committed to the local version. Thus,
insofar as user interaction is concerned, Carmot is qualitatively no
different from any other CRDT system. Unfortunately, we couldn't find
a standardized CRDT system (akin to Quark for MRDTs) for us to do
comparisons against.

### 2022-03-15 abstract-semantics.tex
Incorporating reviewer comments

Fig.~\ref{fig:git-semantics} shows the operational semantics of the
\quark abstract machine. The machine admits the usual version control
operations, namely \rulelabel{Commit}, \rulelabel{Fork},
\rulelabel{Merge}, and \rulelabel{FastFwd}. Operation
\rulelabel{FastFwd} is a special case of merge which simply \emph{fast
forwards} a branch to a later version. A \emph{branch} is a linear
sequence of versions, which intuitively denotes the progressive
evolution of the state on a replica. The latest version of the branch,
called its \emph{head}, denotes the current state of the corresponding
replica. A new head version is created either by an
externally-initiated \emph{commit} (\rulelabel{Commit}) or by merging
the current head with a concurrent or causally-succeeding version from
a different branch (\rulelabel{Merge} and \rulelabel{FastFwd}
respectively). A new branch can be created by \emph{forking off} a
version on an existing branch (\rulelabel{Fork}). Although a version
control system admits more operations (e.g., ``rebase''), we observe
that these four basic actions are sufficient to capture the behavior
of an asynchronously replicated multi-versioned state machine.

The state of \quark abstract machine is a tuple $\Delta = (G,N,C,H,L)$, where:
\begin{itemize}
  \item $G = (V,E)$ is the version history DAG generated by the
    execution of the abstract machine. Vertices ($V$) of the graph are
    the set of all versions that ever existed during the execution.
    Edges ($E$) record the relationships between various versions. In
    particular, there are three kinds of edges corresponding to the
    three basic operations: commit ($\cedge$), merge ($\medge$), fork
    ($\fedge$). Fast forward, being a special case of merge, is also
    denoted by a merge edge ($\medge$). Existence of an edge $v_0
    \rightarrow  v_1$ denotes that versions $v_0$ and $v_1$ are related by
    some operation. For e.g., $v_0 \fedge v_1$ denotes that $v_1$ is a
    new version forked-off from the version $v_0$ using the
    \rulelabel{Fork} rule. As usual, \emph{path} relation is the
    reflexive transitive closure of the edge relation, i.e,
    $\reaches$.
%   \rulelabel{Fork}
%   rule (explained later) dictates that $v_0$ and $v_1$ are different
%   versions belonging to different branches, albeit denoting the

@@

\begin{definition}[Initial State and Version History Graph]
  The graph $G_{\odot} = (\{v_{\odot}\},\emptyset)$ is the initial
  version graph. The state $\Delta_{\odot}$ = $(G_{\odot},\,
  [v_{\odot} \mapsto b_{\odot}],\, [b_{\odot} \mapsto v_{\odot}],\,
  \emptyset)$ is the initial state.
\end{definition}

@@

\paragraph{Rules} The rule \rulelabel{Commit}
(Fig.~\ref{fig:git-semantics}) describes committing a new version $v$
onto the branch $b$ updating its head. The value $n$ for the new
version is assumed to have been provided by whoever has invoked the
commit, e.g., the user. Intuitively, user invokes an RDT operation
(e.g., \C{add($e$)}) on the value of the current version to create the
new value $n$, and thus the new version $v$. A unique commit id $c$ is
assigned to this commit event and added to $C(v)$. The set $C(v)$
also contains all the commit ids from the previous version $H(b)$
since the new value $n$ is assumed to have been derived from the
previous value $N(H(b))$. The edge $H(b) \cedge v$ records this
dependency and also documents the progression of branch $b$. The LCA
map $L$ does not change as the only new edge is between the versions
of the same branch $b$.

@@

\paragraph{Properties} We now formalize the notable properties of the
abstract machine and its executions\footnote{
  The manual proofs of the theorems and their Ivy
  formalization~\cite{ivy} can be found in the supplementary material.
}.

@@

\begin{theorem}[{\bf Progress}]
  \label{thm:progress}
  Every reachable state $\Delta = ((V,E),N,C,H,L)$ of the abstract
  machine is either: 
  \begin{itemize}
    \item A ``quiescent'' state, where: $\forall b_1,b_2 \in
      dom(H).~C(H(b_1)) = C(H(b_2))$, Or
    \item An ``unstuck'' state, where there exist $b_1,b_2 \in dom(H)$
      that satisfy the pre-conditions of \rulelabel{Merge} or
      \rulelabel{FastFwd} rules, i.e., $b_1$ and $b_2$ are mergeable. 
  \end{itemize}
\end{theorem}

@@

>>>>

Fig.~\ref{fig:git-semantics} shows the operational semantics of the
\quark abstract machine. The machine admits the usual version control
operations, namely \rulelabel{Commit}, \rulelabel{Fork},
\rulelabel{Merge}, and \rulelabel{FastFwd}. Operation
\rulelabel{FastFwd} is a special case of merge which simply \emph{fast
forwards} a branch to a later version. A \emph{branch} is a linear
sequence of versions, which intuitively denotes the progressive
evolution of the state on a replica. There is thus a one-to-one
mapping between replicas and branches. The latest version of the
branch, called its \emph{head}, denotes the current state of the
corresponding replica. A new head version is created either by an
externally-initiated \emph{commit} (\rulelabel{Commit}) or by merging
the current head with a concurrent or causally-succeeding version from
a different branch (\rulelabel{Merge} and \rulelabel{FastFwd}
respectively). A new branch can be created by \emph{forking off} a
version on an existing branch (\rulelabel{Fork}). Although a version
control system admits more operations (e.g., ``rebase''), we observe
that these four basic actions are sufficient to capture the behavior
of an asynchronously replicated multi-versioned state machine.

The state of \quark abstract machine is a tuple $\Delta \;=\; $ 
$(G,N,C,H,L)$, where:
\begin{itemize}
  \item $G = (V,E)$ is the version history DAG generated by the
    execution of the abstract machine. Vertices ($V$) of the graph are
    the set of all versions that ever existed during the execution.
    Edges ($E$) record the relationships between various versions. In
    particular, there are three kinds of edges corresponding to the
    three basic operations: commit ($\cedge$), merge ($\medge$), fork
    ($\fedge$). Fast forward, being a special case of merge, is also
    denoted by a merge edge ($\medge$). Existence of an edge $v_0
    \rightarrow  v_1$ denotes that versions $v_0$ and $v_1$ are related by
    some operation. For e.g., $v_0 \fedge v_1$ denotes that $v_1$ is a
    new version forked-off from the version $v_0$ using the
    \rulelabel{Fork} rule. As usual, \emph{path} relation is the
    reflexive transitive closure of the edge relation, i.e,
    $\reaches$.
%   \rulelabel{Fork}
%   rule (explained later) dictates that $v_0$ and $v_1$ are different
%   versions belonging to different branches, albeit denoting the

@@

\begin{definition}[Initial State and Version History Graph]
\label{def:init-state}
  The graph $G_{\odot} = (\{v_{\odot}\},\emptyset)$ is the initial
  version graph. The state $\Delta_{\odot}$ = $(G_{\odot},\,
  [v_{\odot} \mapsto b_{\odot}],\, [b_{\odot} \mapsto v_{\odot}],\,
  \emptyset)$ is the initial state.
\end{definition}

@@

\paragraph{Rules} The rule \rulelabel{Commit}
(Fig.~\ref{fig:git-semantics}) describes committing a new version $v$
onto the branch $b$ updating its head. The value $n$ for the new
version is assumed to have been provided by whoever has invoked the
commit, e.g., the user. Intuitively, user invokes an RDT operation
(e.g., \C{add($e$)}) on the value of the current version to create the
new value $n$, and thus the new version $v$. A unique commit id $i$ is
assigned to this commit event and added to $C(v)$. The set $C(v)$
also contains all the commit ids from the previous version $H(b)$
since the new value $n$ is assumed to have been derived from the
previous value $N(H(b))$. The edge $H(b) \cedge v$ records this
dependency and also documents the progression of branch $b$. The LCA
map $L$ does not change as the only new edge is between the versions
of the same branch $b$.

@@

\paragraph{Properties} We now formalize the notable properties of the
abstract machine and its executions.

@@

\begin{theorem}[{\bf Progress}]
  \label{thm:progress}
  Every reachable state $\Delta = ((V,E),\\N,C,H,L)$ of the abstract
  machine is either: 
  \begin{itemize}
    \item A ``quiescent'' state, where: $\forall b_1,b_2 \in
      dom(H).~C(H(b_1)) = C(H(b_2))$, Or
    \item An ``unstuck'' state, where there exist $b_1,b_2 \in dom(H)$
      that satisfy the pre-conditions of \rulelabel{Merge} or
      \rulelabel{FastFwd} rules, i.e., $b_1$ and $b_2$ are mergeable. 
  \end{itemize}
\end{theorem}

@@

\noindent\paragraph{Ivy formalization} Appendix contains the manual
proofs of the aforementioned theorems. In addition we also formalized
the \quark abstract machine in Ivy~\cite{ivy}, and verified its
properties automatically with help of Z3~\cite{z3}. Ivy is a
multi-modal verification tool, whose primary purpose is to verify the
design and implementations of distributed protocols. The key
differentiating aspect of Ivy is its onus on predictable automation
and decidable reasoning.  Automated proofs in Ivy are restricted to
logical fragments for which the tool is a decision procedure. %Towards
% this end, Ivy checks the verification conditions for syntactic
% confirmity with known decidable fragments, and provides feedback to
% the user if the checks fail. The feedback helps users reorganize their
% formal development or reformulate their queries until the reasoning
% becomes decidable. Ivy uses induction to prove the properties of
% infinite-state systems. The failure to discharge a proof results in a
% counterexample, which is either a real counterexample (i.e., a bug in
% the system) or a counterexample to induction. In the latter case, the
% counterexample serves as a guide for the user to formulate the
% inductive invariants required to discharge the proof. 
Consequently, the creative aspect of proving a theorem in Ivy lies in
discovering decidable abstractions that characterize the system being
reasoned about, and crafting the inductive invariants required to
discharge the automated proof. This is in contrast to theorem provers
such as Coq and Isabelle, and their extensions such as
Sledgehammer~\cite{sledge}, where automation is used on a best-effort
basis to aid interactive theorem proving.

Our formalization of the \quark abstract machine in Ivy follows the
formal development in Fig.~\ref{fig:git-semantics} but uses slightly
different abstractions to obtain decidable reasoning. Maps $C$, $N$,
$H$, and $L$ are modeled as relations instead of functions since
reasoning about quantifier-bound function application is in general
undecidable. $V$ and $E$ are relations by definition. Their initial
values are as specified by $\Delta_{\odot}$ in
Def.~\ref{def:init-state}. The system admits four \emph{actions}
corresponding to the four transition rules in
Fig.~\ref{fig:git-semantics}. Each action ``updates'' the system state
$\Delta = ((V,E),N,C,H,L)$ as defined by the corresponding rule in
Fig.~\ref{fig:git-semantics}.
Commit, version, and branch identifiers are modeled as totally ordered
sets of uninterpreted values from which new values can be chosen
sequentially. This lets us capture these identifiers as monotonically
increasing integers while avoiding undecidable reasoning with
quantifier-bound integer arithmetic.
Lemmas/Theorems~\ref{lem:lca-uniqueness} to~\ref{thm:progress} are
asserted as inductive properties of the \quark abstract machine. Ivy
succesfully verifies each of these properties with help of Z3, thus
certifying the soundness of \quark's meta-theory.

### 2022-03-15 concrete-semantics.tex
Incorporating reviewer comments

\paragraph{Notation and Conventions} We enforce a one-to-one mapping
between replicas and branches by adopting the convention that branch
$b_i$ always corresponds to replica $i$. Concretely this means that
replica $i$ only ever creates new versions on branch $b_i$. Also,
replica $i$ can only update its local copies of $H$ and $B$, i.e.,
$H_i$ and $B_i$. For e.g., $H_i[b_i \mapsto t]$ updates the head of
branch $b_i$ to $t$ on replica $i$.  Since $H_i$ is simply an
abbreviation of $H\;i$ in the formalism, $H_i[b_i \mapsto t]$ actually
expands to $H_i[i \mapsto b_i \mapsto t]$. We exploit this notation in
Fig.~\ref{fig:quark-semantics}. 
% When two replicas, $i$ and $j$, are involved in an update, e.g.,
% \rulelabel{Fork}, we write $H_{\tuplee{i,j}}[b_j \mapsto t]$ to mean
% $H[i \mapsto b_j \mapsto t][j \mapsto b_j \mapsto t]$.

@@

\paragraph{Properties} The Unique LCA
(Lemma~\ref{lem:lca-uniqueness}), Convergence
(Theorem~\ref{thm:convergence}, and Progress
(Theorem~\ref{thm:progress}) properties have been re-defined for the
\quark distributed machine and automatically proven correct in
Ivy~\cite{ivy}.

>>>>

\paragraph{Notation and Conventions} We enforce a one-to-one mapping
between replicas and branches by adopting the convention that branch
$b_i$ always corresponds to replica $i$. Concretely this means that
replica $i$ only ever creates new versions on branch $b_i$. Also,
replica $i$ can only update its local copies of $H$ and $B$, i.e.,
$H_i$ and $B_i$. For e.g., $H_i[b_i \mapsto t]$ updates the head of
branch $b_i$ to $t$ on replica $i$.  Since $H_i$ is simply an
abbreviation of $H\;i$ in the formalism, $H_i[b_i \mapsto t]$ actually
expands to $H_i[i \mapsto b_i \mapsto t]$. We exploit this notation in
Fig.~\ref{fig:quark-semantics}. We let $\delta_{\odot}$ denote the
initial concrete state which is defined on the similar lines as the
initial abstract state (Def.~\ref{def:init-state}).
% When two replicas, $i$ and $j$, are involved in an update, e.g.,
% \rulelabel{Fork}, we write $H_{\tuplee{i,j}}[b_j \mapsto t]$ to mean
% $H[i \mapsto b_j \mapsto t][j \mapsto b_j \mapsto t]$.

@@

\paragraph{Refinement} The following theorem asserts that the \quark
distributed machine refines the abstract machine described in
Sec.~\ref{sec:abstract-sem}:
\begin{theorem}[{\bf Refinement}]
  \label{thm:refinement}
  There exists a refinement relation $R$ relating abstract and
  concrete states such that:
  \begin{itemize}
    \item $R$ relates initial abstract and concrete states, i.e.,
      $R(\Delta_{\odot},\delta_{\odot})$.
    \item Forall $\delta,\delta',\Delta$, if $R(\delta,\Delta)$ and
      $\delta \qstepsto \delta'$, then there exists a $\Delta'$ such
      that $\Delta \stepsto \Delta'$ and $R(\delta',\Delta')$.
  \end{itemize}
\end{theorem}
The proof is done in Ivy. The refinement relation $R$ essentially
relates the version vectors of concrete semantics to the version
histories of abstract semantics. A \emph{happens-before} relation has
been defined as a ghost state in concrete semantics to help us
establish the refinement relation in Ivy.
